---
title: "Comparing the methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing the methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4.2, dpi = 120)
```

```{r setup}
library(trialSize)
```

## Why compare

The plot-size methods disagree by construction, and the published articles
report that disagreement as a finding rather than a nuisance. The MCM optimum is
typically the smallest, the LRP intermediate and the QRP the largest -- an
ordering seen across many crops. `compare_methods()` runs them on one set of
data and tabulates what each recommends, so the spread is visible instead of
inferred from separate calls.

The real question it answers is not *which method gives the largest number* but
*whether the methods disagree by more than their own uncertainty*. With
`bootstrap = TRUE` the answer comes with intervals, and the spread between
methods is often smaller than the imprecision within any one of them.

## From a CV table

Given a data frame with `x` and `cv` columns, the three CV-based methods run:

```{r cvtable}
grid1 <- as.matrix(uniformity_trial[uniformity_trial$trial == "T1",
                                    grep("^col", names(uniformity_trial))])
cv_tab <- calc_cv_shapes(grid1)

cmp <- compare_methods(data.frame(x = cv_tab$x, cv = cv_tab$cv), step = 0.05)
cmp
```

The footer reports the range of $X_o$ across methods as a factor: how many times
larger the largest recommendation is than the smallest.

## From the raw grid

Given a grid of basic experimental units instead of a CV table,
`compare_methods()` builds the CV table on the way with [calc_cv_shapes()], and
the Paranaíba method joins the comparison -- it works on the basic units
directly and so is only available when the grid is supplied:

```{r grid}
compare_methods(grid1, step = 0.05)
```

## Which statistics are comparable

`R2` and `RMSE` are **recomputed here** from the residuals on the original CV
scale, so they mean the same thing for every method even when the fit itself was
weighted -- a weighted fit stores weighted statistics, and comparing those
across methods would compare different quantities.

Two deliberate absences:

- **AIC and BIC** are not shown. The MCM has two parameters against the plateau
  models' three plus a breakpoint, and the breakpoint is not an ordinary
  parameter, so the information criteria are not on a common footing.
- The Paranaíba row has `NA` for `R2` and `RMSE` by nature, not by omission: it
  is a closed form over the basic units with no fitted residuals.

## Weighting

`weights = TRUE` uses the `n` column -- the number of plots each shape yields --
which a grid always provides. It reaches the MCM through that method's Federer
`df` argument, the same weighted least squares by another name. The Paranaíba
row is unaffected, since it never sees the CV table.

```{r weights}
compare_methods(grid1, step = 0.05, weights = TRUE)$summary
```

## Intervals: the point of the exercise

Without intervals the table invites over-reading a difference that may be noise.
`bootstrap = TRUE` adds a confidence interval for each $X_o$, and a
breakpoint-existence p-value for the methods that have one:

```{r bootstrap}
set.seed(1)
compare_methods(grid1, step = 0.05, bootstrap = TRUE, n_boot = 200)
```

The print method reads the intervals for you: if they all share a common
stretch, the methods do not disagree beyond their own uncertainty, and the
choice between them matters less than it looked. If they do not overlap, the
difference is real and worth a decision.

## Several trials at once

A named list of grids compares the methods within each trial:

```{r multi}
grids <- lapply(split(uniformity_trial, uniformity_trial$trial),
                function(d) as.matrix(d[, grep("^col", names(d))]))

compare_methods(grids, step = 0.05)$summary
```

## Choosing the methods

By default every applicable method runs. Restrict them with `methods`; the
Paranaíba method requires a grid and errors if asked for on a CV table:

```{r subset}
compare_methods(grid1, step = 0.05, methods = c("lrp", "qrp"))$summary
```

## Where this fits

`compare_methods()` brings together the methods documented one at a time in
`vignette("lrp")`, `vignette("qrp")`, `vignette("mcm")` and
`vignette("paranaiba")`. Run `vignette("check_trial")` first to know whether the
trial is worth sizing plots against, and `vignette("replicates")` afterwards to
turn the chosen $CV_{Xo}$ into a number of replications.

### References

Cargnelutti Filho, A. et al. (2025). Determinação do tamanho de parcela para
avaliar a massa de parte aérea de grão-de-bico. *Revista Vivências*, 21(43),
499-513, which compares the same three methods and reports 4.81, 7.19 and
10.25 m² for the MCM, LRP and QRP respectively.
