---
title: "The Paranaiba method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Paranaiba method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4.2, dpi = 120)
```

```{r setup}
library(trialSize)
```

## Theory

The other three methods start from a table of CVs, one per planned plot size,
and fit a model to it. The Paranaíba method skips that step entirely: it works
on the **raw grid of basic experimental units** and returns the optimal plot
size in closed form, with no model fitting at all.

The insight is that the whole CV-versus-size curve is driven by how much
neighbouring units resemble one another. If adjacent units are strongly
correlated, grouping them buys little new information and the optimal plot stays
small; if they are independent, larger plots keep paying off. That similarity is
summarized by the first-order spatial autocorrelation $\rho$, and the optimum
follows directly:

$$
X_o = \frac{10 \sqrt[3]{2\,(1 - \rho^{2})\, s^{2}\, m}}{m},
\qquad
CV_{Xo} = \frac{100 \sqrt{(1 - \rho^{2})\, s^{2} / m^{2}}}{\sqrt{X_o}}
$$

where $m$ and $s^2$ are the mean and variance of the BEU values and $\rho$ is
the autocorrelation between adjacent units.

Reading the formula gives the intuition directly: $\rho$ enters only through
$(1 - \rho^2)$, so $X_o$ is largest when $\rho = 0$ (independent units) and
shrinks as $|\rho|$ approaches 1 (strong spatial dependence, in either
direction).

### How rho is estimated

The grid is walked in **serpentine** order, left to right along the first row,
right to left along the second, and so on, turning the two-dimensional grid into
a single sequence in which consecutive entries are physically adjacent. With
$e_i$ the deviations from the mean along that path,

$$
\rho = \frac{\sum_{i=2}^{N} e_i \, e_{i-1}}{\sum_{i=1}^{N} e_i^{2}}
$$

The original method walks in the direction of the **rows**, which is the default
here (`rho_direction = "row"`). `"col"` walks down the columns and `"mean"`
averages both. The choice matters: the two directions can give visibly different
$\rho$, and therefore different optima, especially when the field has a gradient
along one axis.

### References

Original method: Paranaíba, P. F., Ferreira, D. F. & Morais, A. R. (2009).
Tamanho ótimo de parcelas experimentais: proposição de métodos de estimação.
*Revista Brasileira de Biometria*, 27(2), 255-268.

The formulas are validated against published results in the package tests; see
Cargnelutti Filho, A. et al. (2014). Tamanho de parcela e número de repetições
em aveia preta. *Ciência Rural*, 44(10), 1732-1739, where `calc_paranaiba()`
reproduces the published $X_o$ and $CV_{Xo}$ for all nine black-oat trials of
the first evaluation date.

## Data

The example uses the bundled simulated uniformity trial (`?uniformity_trial`):
8 × 12 grids of 1 m² BEU, each holding a biomass-like measurement in g m⁻².

```{r data, message = FALSE}
grid_mat <- function(t)
  unname(as.matrix(uniformity_trial[uniformity_trial$trial == t,
                                    grep("^col", names(uniformity_trial))]))
E1 <- grid_mat("T1")
E2 <- grid_mat("T2")
dim(E1)
```

The matrix must preserve the field layout. Rows and columns are not
interchangeable here: because $\rho$ is directional, a transposed grid gives a
different answer.

## Basic use

```{r fit}
fit <- calc_paranaiba(E1)
fit
```

Both directional estimates are always reported (`rho_row` and `rho_col`), even
though only the one selected by `rho_direction` is used in `rho` and feeds the
formulas. Seeing both is useful: a large gap between them signals a directional
gradient in the field.

```{r summary}
summary(fit)
```

The full per-trial table is in `$summary`, and the grids as supplied are kept in
`$matrices`:

```{r access}
fit$summary[, c("mean", "variance", "CV", "rho", "Xo", "CVxo")]
```

## Direction of the walk

```{r directions}
do.call(rbind, lapply(c("row", "col", "mean"), function(d) {
  f <- calc_paranaiba(E1, rho_direction = d)
  data.frame(direction = d, rho = f$summary$rho,
             Xo = f$summary$Xo, CVxo = f$summary$CVxo)
}))
```

Here both directional estimates are small, so the choice of direction barely
moves $X_o$: $(1 - \rho^2)$ is flat near the origin. In a field with a strong
directional gradient the two would differ more and the choice would matter.

Unless you are deliberately reproducing a study that did otherwise, keep the
default `"row"`, which is the original method.

## Three ways to supply the data

**One matrix**, as above, for a single trial.

**A list of matrices**, for several trials. Names become the trial labels:

```{r list}
res <- calc_paranaiba(list(`Trial 1` = E1, `Trial 2` = E2))
res
```

**A long data frame**, with the measurement column plus row and column indices.
The grid is rebuilt from the indices, not from the row order of the file, which
is much safer when data come out of a spreadsheet:

```{r long}
long <- expand.grid(col = 1:12, row = 1:8)
long$mf <- as.vector(t(E1))

calc_paranaiba(long, value = "mf", row_id = "row", col_id = "col")$summary$Xo
```

With a `trial` column, one estimate is produced per trial:

```{r long-trial}
long2 <- rbind(
  transform(long, trial = "Trial 1"),
  transform(expand.grid(col = 1:12, row = 1:8), mf = as.vector(t(E2)),
            trial = "Trial 2")
)

calc_paranaiba(long2, value = "mf", row_id = "row", col_id = "col",
               trial = "trial")$summary[, c("trial", "rho", "Xo", "CVxo")]
```

Missing cells are an error rather than a silent gap, and `n_row` / `n_col` can
be passed to assert the expected grid dimensions.

## Plot

```{r plot}
plot(res, title = "Paranaiba: optimal plot size")
```

The dashed line is the mean across trials. Use `y_var = "CVxo"` to show the CV
at the optimum instead:

```{r plot-cv}
plot(res, y_var = "CVxo", title = "CV at the optimal plot size")
```

```{r save, eval = FALSE}
plot(res, title = "Paranaiba",
     save = TRUE, file = "paranaiba.tiff", format = "tiff", dpi = 300)
```

## When results are flagged invalid

A trial with zero variance (every unit identical) or a zero mean cannot produce
an estimate: $\rho$ is undefined and the formulas divide by the mean. Those
trials get `valid = FALSE` and `NA` results, with a warning naming the trial,
rather than propagating `NaN` quietly into a summary table.

## How it compares

The Paranaíba estimate uses information the CV-based methods discard, namely
where each unit sits in the field, and it needs no model fitting or convergence.
The trade-off is that it needs the raw grid, which the CV-based methods do not.

If you have the raw grid, running both is worthwhile: agreement between a
spatial estimate and a CV-curve estimate is reassuring, and disagreement usually
points at a directional gradient worth investigating before designing the
experiment.

See `vignette("lrp")`, `vignette("qrp")` and `vignette("mcm")` for the CV-based
methods, and `vignette("replicates")` for turning $CV_{Xo}$ into a number of
replications.
